%Acknowledgments

\small\normalsize
\hbox{\ }
 
\vspace{-.65in}


\chapter*{\LARGE \begin{center}Acknowledgements\end{center}} % this is a bit hackish
\addcontentsline{toc}{chapter}{Acknowledgements}
\vspace{40pt}


I am deeply indebted to Philip Resnik for guiding me through this process.  This dissertation is a testament to his faith in me, which was often greater than my own.  I would not have found my own path if he had not given me the freedom to get lost, and direction when I stayed lost for too long.  His enthusiasm and encouragement were crucial to my success, and I can only hope that I one day inspire someone else as much as he has inspired me.

I am also grateful for the guidance of my committee---David Chiang, Bonnie Dorr, Doug Oard, and David Fushman.  
Much of the work described herein was influenced in one way or another by my interactions with David Chiang.  In 2005, I asked him whether the suffix array trick would work with hierarchical models, and he told me he didn't think so.  It's about the only time I've ever known him to be wrong about anything, and even that took over two years and an enormous amount of work to prove.  David also gave me very constructive early feedback that shaped the survey chapter, and it is not hyperbole to say that I learned much of what I know about statistical translation just from reading his elegant code.
Bonnie Dorr gave me tremendously detailed and helpful feedback on several drafts of the dissertation, and Doug Oard directly challenged me to broaden its scope and vision.  The final result is much stronger than it would have been without them.
Finally, David Fushman gamely served on my committee and engaged with me throughout this process even though his own work is very far afield from mine.  Together with Philip, they held this dissertation to a high standard.  Any errors that remain are my own.

Several people in the Department of Computer Science and UMIACS helped along the way.  I am grateful to Bill Gasarch for serving on my proposal committee, but more importantly for his unvarnished advice.  Mihai Pop graciously allowed me to audit and actively engage with his bioinformatics seminar, where I found most of the tools that I needed to solve David's challenge.  Fritz McCall went above and beyond the call of the duty assisting me through my forays into network programming.

I learned most of what I know about natural language processing from the present and past students, postdoctoral researchers, and visitors who have been my colleagues in the CLIP lab.  Many ideas were developed and refined in discussions with
Fazil Ayan,
Dina Demner-Fushman,
Mona Diab,
Nizar Habash,
Tim Hunter,
Okan Kolak,
Nitin Madnani,
Smaranda Muresan,
Christof Monz,
Mike Nossal,
Michael Subotin,
Davic Zajic, and
Dan Zeman.
I owe special thanks to Chris Dyer for all of the brainstorming, critiquing, and commiserating during the course of writing this dissertation.  Most especially, I thank Rebecca Hwa for being my mentor and my friend.

Outside of the lab, I thank Bill Woods for my earliest exposure to NLP research, and I'm enormously grateful to Philipp Koehn---first for offering me a fantastic job that I'm very excited about, and then for waiting on me while I completed this obligation.

I had a lot of fun in grad school.  This I owe to the friends I spent time with outside of the lab:
Laura Bright,
Gene Chipman,
Ric Crabbe,
Shaun Gittens,
Debbie Heisler,
Omer Horvitz,
Dave Hovemeyer,
Doc Hunter,
Aram Khalili,
Jordan Landes,
Jeremy Manson,
Jan Neumann,
Brian Postow,
Reiner Schulz,
Jaime Spacco,
Kate Swope, and
Chadd Williams.
Thank you all for keeping me sane.

My family has been supportive of me throughout this endeavor, even though it seemed prolonged and bewildering when beheld from afar.  I am thankful to Lisa and Josh Friedman, and my parents Richard and Edith Lopez.

Most importantly, Jennifer Baugher has filled my days, nights, and dreams with love throughout my time here.  She has been with me for every moment of triumph and despair, through countless adventures and many bottles of wine.  I know I could accomplish anything when she is with me.  Thank you for everything, for ever.

\newpage
~
\vspace{1.5in}

This research was supported in part by NSA Contract RD-02-5700, ONR MURI Contract FCPO.810548265 and the GALE program of the Defense Advanced Research Projects Agency, Contract No. HR0011-06-2-001. 
Any opinions, findings, conclusions or recommendations expressed in this work are those of the author and do not necessarily reflect the view of DARPA.  
Additional support was provided by the EuroMatrix project funded by the European Commission (6th Framework Programme).


